---
title: 'Machine Learning: Lab 1a'
author: "Scout Leonard"
date: "1/3/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## 2.1 Install Packages

```{r}
# load packages, installing if missing
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  dismo, dplyr, DT, ggplot2, here, htmltools, leaflet, mapview, purrr, raster, readr, rgbif, rgdal, rJava, sdmpredictors, sf, spocc, tidyr, geojsonio)
select <- dplyr::select # overwrite raster::select

# set random seed for reproducibility
set.seed(42)

# directory to store data
dir_data <- here("data/sdm")
dir.create(dir_data, showWarnings = F, recursive = T)
```

## 2.2 Choose a Species

I chose the **Atlantic Horseshoe Crab**, _Limulus polyphemus_.

## 2.3 Get Species Observations

```{r}
obs_csv <- file.path(dir_data, "obs.csv")
obs_geo <- file.path(dir_data, "obs.geojson")

# get species occurrence data from GBIF with coordinates
(res <- spocc::occ(
  query = 'Limulus polyphemus', 
  from = 'gbif', has_coords = T,
  limit = 10000))
```

```{r}
# extract data frame from result
df <- res$gbif$data[[1]] 
nrow(df) # number of rows
```

```{r}
#remove odd observations
df <- df %>% 
  filter(!stateProvince %in% c("Indiana", "California")) %>% #california and indiana outlier
  filter(key != 3398866235) %>% #south carolina northerly outlier
  filter(key != 3302425640) #virginia westerly outlier

# convert to points of observation from lon/lat columns in data frame
obs <- df %>% 
  sf::st_as_sf(
    coords = c("longitude", "latitude"),
    crs = st_crs(4326))

readr::write_csv(df, obs_csv)
geojsonio::geojson_write(obs, file = "data/sdm/obs.geojson")

# show points on map
mapview::mapview(obs$geometry, map.types = "Esri.OceanBasemap")
```

**Question 1: How many observations total are in GBIF for your species?**

```{r}
res
```

There are 8,561 total observations in GBIF for _Limulus polyphemus_.

**Question 2: Do you see any odd observations, like marine species on land or vice versa? If so, please see the Data Cleaning and explain what you did to fix or remove these points.**

There were 3 observations that were extremely inland for a coastal marine species and one observation in the Pacific Ocean. I filtered them from my dataframe by clicking the outliers, finding the row they appear in, and retrieving their identifier from the `key` column. I filter based on that for the 4 odd observations and was able to return a more reasonable map based on what I know about Atlantic Horseshoe Crabs. 

## 2.4 Get Environmental Data 

### 2.4.1 Presence

```{r}
dir_env <- file.path(dir_data, "env")

# set a default data directory
options(sdmpredictors_datadir = dir_env)

# choosing terrestrial and marine - because this species is coastal
env_datasets <- sdmpredictors::list_datasets(terrestrial = TRUE, marine = TRUE)

# show table of datasets
env_datasets %>% 
  select(dataset_code, description, citation) %>% 
  DT::datatable()
```

```{r}
# choose datasets for a vector
env_datasets_vec <- c("WorldClim", "ENVIREM")

# get layers
env_layers <- sdmpredictors::list_layers(env_datasets_vec)
DT::datatable(env_layers)
```

```{r}
# choose layers after some inspection and perhaps consulting literature
env_layers_vec <- c("WC_alt", "WC_bio1", "WC_bio2", "ER_tri", "ER_topoWet")

# get layers
env_stack <- load_layers(env_layers_vec)

# interactive plot layers, hiding all but first (select others)
#mapview(env_stack, hide = T)
plot(env_stack, nc=2)
```

```{r}
obs_hull_geo  <- file.path(dir_data, "obs_hull.geojson")
env_stack_grd <- file.path(dir_data, "env_stack.grd")

if (!file.exists(obs_hull_geo) | redo){
  # make convex hull around points of observation
  obs_hull <- sf::st_convex_hull(st_union(obs))
  
  # save obs hull
  write_sf(obs_hull, obs_hull_geo)
}
obs_hull <- read_sf(obs_hull_geo)

# show points on map
mapview(
  list(obs, obs_hull))
```

```{r}
if (!file.exists(env_stack_grd) | redo){
  obs_hull_sp <- sf::as_Spatial(obs_hull)
  env_stack <- raster::mask(env_stack, obs_hull_sp) %>% 
    raster::crop(extent(obs_hull_sp))
  writeRaster(env_stack, env_stack_grd, overwrite=T)  
}
env_stack <- stack(env_stack_grd)

# show map
# mapview(obs) + 
#   mapview(env_stack, hide = T) # makes html too big for Github
plot(env_stack, nc=2)
```


```{r}
# show points on map
mapview(list(obs$geometry, obs_hull))
```

### 2.4.2 Pseudo-Absence 

```{r}
absence_geo <- file.path(dir_data, "absence.geojson")
pts_geo     <- file.path(dir_data, "pts.geojson")
pts_env_csv <- file.path(dir_data, "pts_env.csv")

# get raster count of observations
r_obs <- rasterize(
  sf::as_Spatial(obs), env_stack[[1]], field=1, fun='count')

mapview(obs$geometry) + 
  mapview(r_obs)
```

```{r}
# create mask for 
r_mask <- mask(env_stack[[1]] > -Inf, r_obs, inverse=T)

absence <- dismo::randomPoints(r_mask, nrow(obs)) %>% 
  as_tibble() %>% 
  st_as_sf(coords = c("x", "y"), crs = 4326)

mapview(obs$geometry, col.regions = "green") + 
  mapview(absence, col.regions = "gray")
```

```{r}
# combine presence and absence into single set of labeled points 
pts <- rbind(
  obs %>% 
    mutate(
      present = 1) %>% 
    select(present),
  absence %>% 
    mutate(
      present = 0)) %>% 
  mutate(
    ID = 1:n()) %>% 
  relocate(ID)
write_sf(pts, pts_geo)

# extract raster values for points
pts_env <- raster::extract(env_stack, as_Spatial(pts), df=TRUE) %>% 
  tibble() %>% 
  # join present and geometry columns to raster value results for points
  left_join(
    pts %>% 
      select(ID, present),
    by = "ID") %>% 
  relocate(present, .after = ID) %>% 
  # extract lon, lat as single columns
  mutate(
    #present = factor(present),
    lon = st_coordinates(geometry)[,1],
    lat = st_coordinates(geometry)[,2]) %>% 
  select(-geometry)

write_csv(pts_env, pts_env_csv)
```

```{r}
pts_env %>% 
  select(-ID) %>% 
  DT::datatable()
```

## 2.5 Term Plots 

```{r}
pts_env %>% 
  select(-ID) %>% 
  mutate(
    present = factor(present)) %>% 
  pivot_longer(-present) %>% 
  ggplot() +
  geom_density(aes(x = value, fill = present)) + 
  scale_fill_manual(values = alpha(c("gray", "green"), 0.5)) +
  scale_x_continuous(expand=c(0,0)) +
  scale_y_continuous(expand=c(0,0)) +
  theme_bw() + 
  facet_wrap(~name, scales = "free") +
  theme(
    legend.position = c(1, 0),
    legend.justification = c(1, 0))
```
